Flow-Based Cost Allocation in Linear Optimized Power Systems are non-unique but localizable

### 2020-08-11 paper/flow_based.tex
fix .gitignore
update numerical example

% network
\newcommand{\incidence}[1][n]{K_{#1,\ell}}
\newcommand{\incidencegenerator}[1][n]{K_{#1,s}}
\newcommand{\incidencestorage}[1][n]{K_{#1,r}}
\newcommand{\ptdf}[1][n]{H_{\ell,#1}}
\newcommand{\cycle}{C_{\ell,c}}
\newcommand{\impedance}{x_\ell}
\newcommand{\cycleprice}{\lambda_{c,t}}

% allocation quantities
\newcommand{\allocategeneration}[1][s, n]{A_{#1,t}}
\newcommand{\allocatepeer}[1][m \rightarrow n]{A_{#1,t}}
\newcommand{\allocateflow}[1][n]{A_{\ell,#1,t}}
\newcommand{\allocatetransaction}[1][s \rightarrow n]{A_{#1,\ell,t}}
\newcommand{\allocatestate}[1][i, n]{A_{#1,t}}
\newcommand{\allocatecapexgeneration}[1][n]{\mathcal{C}^{G}_{#1,t}}
\newcommand{\allocatecapexflow}[1][n]{\mathcal{C}^{F}_{#1,t}}
\newcommand{\allocateopex}[1][n]{\mathcal{O}_{#1,t}}
\newcommand{\allocateemissioncost}[1][n]{\mathcal{E}_{#1,t}}

@@

% where $c_{i,t}$ is either a fixed operational price or a composition of shadowprices
Following the implications of \cref{eq:lmp,eq:total_cost_reduced},  we define the cost $\cost_{n\rightarrow i,t}$ that consumers at bus $n$ have pay to asset $i$ at time $t$ as  
\begin{align}
    \cost_{n \rightarrow i,t} = \costfactor \,\pdv{\state}{\demand} \demand
\end{align}
Therefore, $\cost_{n\rightarrow i,t}$ indicates the contribution of $\demand$ to the cost term $\cost_{i}$. It heavily relies on the derivative of the operational state with respect to the nodal demand $\partial \state / \partial \demand$. Without further specifying, we define the derivative multiplied with the nodal demand as 
\begin{align}
    \allocatestate = \pdv{\state}{\demand} \demand
    \label{eq:allocate_state}
\end{align}    
This quantity may be interpreted as the amount of power that asset $i$ supplies demand $\demand$ with.
From the natural fact that the sum of all contributions must return the cost term, 
\begin{align}
    \cost_i = \sum_{n,t} \cost_{n \rightarrow i, t}
\end{align}
it follows that $\allocatestate$ must fulfill
\begin{align}
    \state = \sum_n \allocatestate
    \label{eq:allocate_state_constraint}
\end{align}
Finally, the total contribution from node $n$ at time $t$ to the cost term $\cost$ amounts 
\begin{align}
    \cost_{n,t} = \sum_i \cost_{n \rightarrow i, t}
\end{align}
\end{subequations}

@@

The nominal capacity $\capacitygeneration$ constrains the generation $\generation$ in the form of 
\begin{align}
\generation - \generationpotential \capacitygeneration  &\le 0 \resultsin{\muuppergeneration} \Forall{s,t} 
\label[constraint]{eq:upper_generation_capacity_constraint}\\ 
- \generation &\le 0 \resultsin{\mulowergeneration} \Forall{s,t} 
\label[constraint]{eq:lower_generation_capacity_constraint}
\end{align}
where $\generationpotential \in \left[ 0,1\right]$ is the capacity factor for renewable generators. At a cost-optimum, these two constraints yield the shadow prices $\muuppergeneration$ and $\mulowergeneration$.  As shown in \cite{brown_decreasing_2020} and in detailed in \cref{sec:zero_profit_generation}, over the whole time span, the CAPEX for generator $s$ is payed back by the production $\generation$ times the shadow price $\muuppergeneration$, 
\begin{align}
 \capexgeneration_s = \capitalpricegeneration \capacitygeneration = \sum_t \muuppergeneration \,  \generation 
 \label{eq:no_profit_capex_generation}
\end{align}
This representation connects the CAPEX with the operational state of generator $s$ and matches the needed form in \cref{eq:cost_decomposition} for using the allocation scheme. The resulting the CAPEX allocation is given by
\begin{align}
 \allocatecapexgeneration[n \rightarrow s] = \muuppergeneration \, \allocategeneration
 \label{eq:allocate_capexGeneration_detailed}
\end{align}
How does this allocation behave? According to the polluter pays principle, it differentiates between consumers who are `responsible` for investments and those who are not. If $\muuppergeneration$ (in literature often denoted as the Quality of Supply) is bigger than zero, the upper Capacity \cref{eq:upper_generation_capacity_constraint} is binding. Thus it is these times steps which push investments in $\capacitygeneration$. If $\muuppergeneration = 0$, the generation $\generation$ is not bound and investments are not necessary. 
When summing over all CAPEX payments to generator $s$ in \cref{eq:allocate_capexGeneration_detailed} and using \cref{eq:allocate_state_constraint,eq:no_profit_capex_generation}, we see that each generator retrieves exactly the cost that were spent to build the capacity $\capacitygeneration$.

@@

The transmission capacity $\capacityflow$ limits the flow $\flow$ in both directions,
\begin{align}
\flow - \capacityflow &\le 0 \resultsin{\muupperflow} \Forall{\ell,t} 
\label[constraint]{eq:upper_flow_capacity_constraint} \\
- \flow - \capacityflow &\le 0 \resultsin{\mulowerflow} \Forall{\ell,t} 
\label[constraint]{eq:lower_flow_capacity_constraint}
\end{align}
which yield the shadow prices $\muupperflow$ and $\mulowerflow$. Again, we use the result of \cite{brown_decreasing_2020} (for details see \cref{sec:zero_profit_flow}) which derives that over the whole time span, the investment in line $\ell$ is payed back by the shadow prices times the flow 
\begin{align}
\capexflow_\ell = \capitalpriceflow \capacityflow = \sum_{t} \left( \muupperflow - \mulowerflow \right)  \flow 
\label{eq:no_profit_capex_flow}
\end{align}
Again, we follow the scheme in \cref{eq:general_scheme} which  defines the CAPEX allocation as 
\begin{align}
    \allocatecapexflow[n \rightarrow \ell] &=  
   \left( \muupperflow - \mulowerflow\right) \, \allocateflow
   \label{eq:allocate_capexFlow_detailed}
\end{align}
The quantity 
\begin{align}
 \allocateflow =  \pdv{\flow}{\demand}\, \demand
\end{align}
can be interpreted as the flow that the demand at $\left(n,t\right)$ causes on line $\ell$.
% According to \cref{eq:allocate_state_constraint} it has to fulfill 
% \begin{align}
%  \flow = \sum_{n} \allocateflow
% \label{eq:allocate_flow_constraint}
% \end{align}
The  shadowprices $\muupperflow$ and $\mulowerflow$ again can be seen as a measure for necessity of transmission investments at $\ell$ at time $t$. Hence, the definition of $\allocatecapexflow[n \rightarrow \ell]$ states that consumers, which retrieve power flowing on congested lines, yielding a bound \cref{eq:upper_flow_capacity_constraint} or \eqref{eq:lower_flow_capacity_constraint}, pay compensations for the resulting investments at $\ell$. Again the sum of all CAPEX payments to line $\ell$ equals the total CAPEX spent. This is seen when summing \cref{eq:allocate_capexFlow_detailed} over all buses and time steps and using \cref{eq:allocate_state_constraint,eq:no_profit_capex_flow}

@@

Given a fix price for emissions in CO$_2$ equivalents $\emissionprice$, the cost term for emission adds up to 
\begin{align}
 \emissioncost = \emissionprice \, \sum_s  \emission \, \generation
\end{align}
where $\emission$ denotes the emission factor in tonne-CO$_2$ per \megawatthour for generator $s$.
We use the same set of equation as given through \cref{eq:general_scheme,eq:opex_allocation} but update the cost factor to $\costfactor \Rightarrow \emissionprice \, \emission$. For consumers at bus $n$ at time $t$ this leads to the emission cost of 
\begin{align}
 \allocateemissioncost[n \rightarrow s] = \emissionprice \, \emission \, \allocategeneration
\end{align}
associated with generator $s$.

@@

The presented cost allocation suits for any type of topology and network setup. But so far, the question of how $\allocatestate$ for generators $s$, lines $\ell$ and storages $r$ are defined was left open. We recap that all rely on derivatives of $\generation$, $\flow$ and $\storagedispatch$  with respect to the nodal demand $\demand$ (see \cref{eq:allocate_state_constraint}). \\
% 
In fact, it turns out that ones, the peer-to-peer allocations $\allocatepeer$ are defined, the power flow allocations $\allocateflow$ are determined by the linear power flow equations.
In order to show this, let $\ptdf$ denote Power Transfer Distribution Factors (PTDF) giving the changes in the flow on line $\ell$ for one unit (typically one MW) of net power production at bus $n$.  Further let $\incidencegenerator$ be 1 if generator $s$ is attached to bus $n$ and zero otherwise. The linear power flow equation can be written as 
\begin{align}
 \flow  = \sum_m \ptdf[m] \left( \nodalgeneration[m] - \demand[m] \right)  
\end{align}
where the nodal generation $\nodalgeneration[m] = \sum_s \incidencegenerator[m] \, \generation $ combines the production of all generators attached to bus $m$. Taking the derivative with respect to the demand, leads to 
\begin{align}
 \allocateflow = \pdv{\flow}{\demand} \demand = \sum_m \ptdf[m] \left( \allocatepeer  - \delta_{n,m} \demand \right) 
 \label{eq:allocate_peer_to_allocate_flow}
\end{align}
where $\allocatepeer = \sum_s \incidencegenerator \, \allocategeneration$ combines all power allocations of the source $m$. \Cref{eq:allocate_peer_to_allocate_flow} shows that the peer-to-peer allocation $\allocategeneration$ directly determines the flow allocation on line $\ell$.

This leaves us with only solving \cref{eq:allocate_peer_constraint} for $\allocategeneration$. In fact, we have to answer the question of how much power supply does generator $s$ deliver to bus $n$. Unfortunately, the solution is non-unique and requires further assumptions. Established flow allocation schemes approach this problem from different directions. Principally two options exist \textit{what} is allocated 
% 
\begin{enumerate}
\item gross power injections \label{gross}
\item net power injections \label{net}
\end{enumerate}
% 
Further it is important \textit{what assumptions} define the allocation, \ie what method is used to define the pairs of sources and sinks. The three suitable approaches we present here are
% 
\begin{enumerate}[label=\alph*., ref=\alph*]
\item Equivalent Bilateral Exchanges (EBE) \cite{galiana_transmission_2003} which assumes
that every producer supplies every consumer proportional to its share in the total consumption. \label{ebe} 
\item Average Participation (AP) \cite{bialek_tracing_1996,achayuthakan_electricity_2010-1} which traces the flow from producer to consumer following the law of proportional sharing. \label{ap}
\item Flow Based Market Coupling (FBMC) which uses zonal PTDF for allocating power within predefined regions. The interregional exchange is only allocating net power decifit or excess of the regions. \label{fbmc}
\end{enumerate}
% 
We show the mathematical formulation for all combinations \ref{ebe}\ref{gross} - \ref{fbmc}\ref{net} in \cref{sec:gross_ebe,sec:net_ebe,sec:gross_ap,sec:net_ap}.
% 
Principally, type \ref{net} leads to less P2P trades then type \ref{gross} as power from a bus $n$ with $\nodalgeneration \le \nodaldemand$ is not assigned to other buses, only to $n$. 
Further, as literature has often pointed out, the EBE principal \ref{ebe} does not suit for large networks where remote buses would interconnect in the same way as buses in close vicinity \cite{gil_multiarea_2005}. The AP based type \ref{ap} tackles this problem by restricting P2P trades to those which are traceable when applying the proportional sharing principal. Therefore $\allocategeneration$ denotes that part of power produced by bus $m$ which, when only following in the direction of $\flow$, ends up at bus $n$. Type \ref{fbmc} further allows to control the regions or market zones which are netted out in a first calculation. If in a region $R$ the generation undercuts the demand, $\sum_{n \in R} \generation \le \sum_{n \in R} \demand$, none of the inner-regional generation is assigned to other regions.

@@

In the following we showcase a two bus system with one optimized time step and its resulting allocated payments, illustrated in \cref{fig:example_network}. \\
% 
% 
\begin{figure*}[t]
\centering
\includegraphics[width=\linewidth]{example_network.png}
\caption{Illustrative example of a 2-bus network with one optimized time step. Fixed prices and constraining values are given in the left box for each bus and the transmission line. Optimized values are given in the right boxes. Bus~1 has a cheaper operational price $o$, capital prices are the same for both. As both generator capacities are constraint to 100~MW, the optimization also deploys the generator at bus~2. The resulting electricity prices $\lambda$ are then a composition of all prices for operation and capital investments. \vspace{-10pt}}
\label{fig:example_network}
\end{figure*}
% 
% 
The two buses are connected via one transmission line, each has one generator. Whereas the generator at bus~1 has an operational price of 50 \euro/\megawatthour, the generator at bus~2 has a higher operational price of 200~\euro/\megawatthour. For both the CAPEX rate is set to 500~\euro/MW and the maxmimal capacity is limited to $\capacitygenerationupper$~=~100~MW. The transmission line has a CAPEX rate of 100~\euro/MW and no upper capacity limit. With a demand of 60~MW at bus~1 and 90~MW at bus~2, the optimization expands the cheaper generator at bus~1 to its full limit of 100~MW. The 40~MW excess power not consumed at bus~1, flow to bus~2 where the generator is built with only 50~MW.  
% 
\begin{figure}[h!]
    \begin{subfigure}[c]{\linewidth}
    \includegraphics[width=\linewidth]{example_allocation_bus1.png}
    \vspace{-40pt}
    \subcaption{}
    \label{fig:example_allocation_bus1}
    \end{subfigure}
    \begin{subfigure}[c]{\linewidth}
    \includegraphics[width=\linewidth]{example_allocation_bus2.png}
    \vspace{-40pt}
    \subcaption{}
    \label{fig:example_allocation_bus2}
    \end{subfigure}
    \caption{Power allocations for bus~1 (a) and bus~2 (b) of the example network in \cref{fig:example_network}. Bus~1 retrieves 40~MW from itself and 20~MW from bus~2. The latter in turn retrieves 60~MW from bus~1 and self supplies 30~MW.
    The sum of both net flows equals the resulting flow of $f=40$~MW. \vspace{-20pt}}
    \label{fig:example_allocation}
\end{figure}
% 
% 
% 
\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{example_payoff.png}
\caption{Full P2P cost allocation for the example setup shown in \cref{fig:example_network}. The payments are derived on the basis of \cref{eq:allocate_opexGeneration_detailed,eq:allocate_capexGeneration_detailed,eq:allocate_capexFlow_detailed}. Consumers at bus $n$ have to pay each generator proportional to their consumption. As we only consider one time step the proportionality applies for OPEX $\mathcal{O}_n$ and CAPEX $\mathcal{C}_n$. As bus~1 induces a relieving flow an line~1 and therefore ``prevents`` further tranmission expansion, it is rewarded proportional to the relief.}
\label{fig:example_payoff}
\end{figure}
% 
% 
\subsubsection*{Allocating Gross Power using EBE (\ref{gross}\ref{ebe}) }

\Cref{fig:example_allocation} shows the allocated transactions on basis of equvalently traded gross power (type \ref{gross}\ref{ebe}) for both buses 1~\&~2 separately. The resulting P2P payments are given in \cref{fig:example_payoff}.
The upper graph \cref{fig:example_allocation_bus1} shows that $A_{1,1}=40$~MW at bus~1 are self-sustained. Consumers at bus~1 consequently pay 2k~\euro~OPEX and 22k~\euro~CAPEX to the generator at bus~1. The remaining 20~MW come from bus~2 and induce a subflow on line 1 of $A_{2 \rightarrow 1, 1} = -20$. As this flow is in contrary direction to the total flow, it is relieving the transmission system. This translates to a congestion reward for consumers at bus~1 of $c_{\ell=1} A_{2 \rightarrow 1, 1}$~=~2k~\euro\, which is exactly the cost that had to be spent on the transmission system if bus~1 didn't induce a relieving flow, see again \cref{fig:example_payoff}.

@@

\begin{align}
    \capitalpricegeneration \capacitygeneration - \remainingcost_s = \sum_{n,t} \allocatecapexgeneration[n \rightarrow s] \Forall{s \in S}
\end{align}

Doing likewise with a subset $L$ of transmission lines,
\begin{align}
\capacityflow \ge \capacityflowLower \resultsin{\mulowerflownom} \hpad \forall{\ell \in \textit{L}} \\
\capacityflow \le \capacityflowUpper \resultsin{\muupperflownom} \hpad \forall{\ell \in \textit{L}}
\end{align}
results in a shifted zero-profit rule for transmission line in the form 
\begin{align}
    \capitalpriceflow \, \capacityflow - \remainingcost_\ell = \sum_{n,t}\allocatecapexflow[n \rightarrow \ell] \Forall{\ell \in L}
\end{align}
where $\remainingcost_\ell$ is given by 
\begin{align}
    \remainingcost_\ell = \left(\mulowerflownom - \muupperflownom \right) \, \capacityflow
    \Forall{\ell \in L}
\end{align}

>>>>

% network
\newcommand{\incidence}[1][n]{K_{#1,\ell}}
\newcommand{\incidencegenerator}[1][n]{K_{#1,s}}
\newcommand{\incidencestorage}[1][n]{K_{#1,r}}
\newcommand{\incidenceasset}[1][n]{K_{#1,i}}
\newcommand{\ptdf}[1][n]{H_{\ell,#1}}
\newcommand{\cycle}{C_{\ell,c}}
\newcommand{\impedance}{x_\ell}
\newcommand{\cycleprice}{\lambda_{c,t}}

% allocation quantities
\newcommand{\allocategeneration}[1][s, n]{A_{#1,t}}
\newcommand{\allocatestoragedispatch}[1][r, n]{A_{#1,t}}
\newcommand{\allocatepeer}[1][m \rightarrow n]{A_{#1,t}}
\newcommand{\allocateflow}[1][n]{A_{\ell,#1,t}}
\newcommand{\allocatetransaction}[1][s \rightarrow n]{A_{#1,\ell,t}}
\newcommand{\allocatestate}[1][i, n]{A_{#1,t}}
\newcommand{\allocatecost}[1][n \rightarrow i]{\mathcal{C}_{#1, t}}
\newcommand{\allocatecapexgeneration}[1][n \rightarrow s]{\mathcal{C}^{G}_{#1,t}}
\newcommand{\allocatecapexflow}[1][n \rightarrow \ell]{\mathcal{C}^{F}_{#1,t}}
\newcommand{\allocateopex}[1][n \rightarrow s]{\mathcal{O}_{#1,t}}
\newcommand{\allocateemissioncost}[1][n \rightarrow s]{\mathcal{E}_{#1,t}}

@@

% where $c_{i,t}$ is either a fixed operational price or a composition of shadowprices
Following the implications of \cref{eq:lmp,eq:total_cost_reduced},  we define the cost $\allocatecost$ that consumers at bus $n$ have pay to asset $i$ at time $t$ as  
\begin{align}
    \allocatecost = \costfactor \,\pdv{\state}{\demand} \demand
\end{align}
Therefore, $\allocatecost$ indicates the contribution of $\demand$ to the cost term $\cost_{i}$. It heavily relies on the derivative of the operational state with respect to the nodal demand $\partial \state / \partial \demand$. Without further specifying, we define the derivative multiplied with the nodal demand as 
\begin{align}
    \allocatestate = \pdv{\state}{\demand} \demand
    \label{eq:allocate_state}
\end{align}    
This quantity may be interpreted as the amount of power that asset $i$ supplies demand $\demand$ with.
From the natural fact that the sum of all contributions must return the cost term, 
\begin{align}
    \cost_i = \sum_{n,t} \allocatecost
\end{align}
it follows that $\allocatestate$ must fulfill
\begin{align}
    \state = \sum_n \allocatestate
    \label{eq:allocate_state_constraint}
\end{align}
Finally, the total contribution from node $n$ at time $t$ to the cost term $\cost$ amounts 
\begin{align}
    \cost_{n,t} = \sum_i \allocatecost
\end{align}
\end{subequations}

@@

The nominal capacity $\capacitygeneration$ constrains the generation $\generation$ in the form of 
\begin{align}
\generation - \generationpotential \capacitygeneration  &\le 0 \resultsin{\muuppergeneration} \Forall{s,t} 
\label[constraint]{eq:upper_generation_capacity_constraint}\\ 
- \generation &\le 0 \resultsin{\mulowergeneration} \Forall{s,t} 
\label[constraint]{eq:lower_generation_capacity_constraint}
\end{align}
where $\generationpotential \in \left[ 0,1\right]$ is the capacity factor for renewable generators. At a cost-optimum, these two constraints yield the shadow prices $\muuppergeneration$ and $\mulowergeneration$.  As shown in \cite{brown_decreasing_2020} and in detailed in \cref{sec:zero_profit_generation}, over the whole time span, the CAPEX for generator $s$ is payed back by the production $\generation$ times the shadow price $\muuppergeneration$, 
\begin{align}
 \capexgeneration_s = \capitalpricegeneration \capacitygeneration = \sum_t \muuppergeneration \,  \generation 
 \label{eq:no_profit_capex_generation}
\end{align}
This representation connects the CAPEX with the operational state of generator $s$ and matches the needed form in \cref{eq:cost_decomposition} for using the allocation scheme. The resulting the CAPEX allocation is given by
\begin{align}
 \allocatecapexgeneration = \muuppergeneration \, \allocategeneration
 \label{eq:allocate_capexGeneration_detailed}
\end{align}
How does this allocation behave? According to the polluter pays principle, it differentiates between consumers who are `responsible` for investments and those who are not. If $\muuppergeneration$ (in literature often denoted as the Quality of Supply) is bigger than zero, the upper Capacity \cref{eq:upper_generation_capacity_constraint} is binding. Thus it is these times steps which push investments in $\capacitygeneration$. If $\muuppergeneration = 0$, the generation $\generation$ is not bound and investments are not necessary. 
When summing over all CAPEX payments to generator $s$ in \cref{eq:allocate_capexGeneration_detailed} and using \cref{eq:allocate_state_constraint,eq:no_profit_capex_generation}, we see that each generator retrieves exactly the cost that were spent to build the capacity $\capacitygeneration$.

@@

The transmission capacity $\capacityflow$ limits the flow $\flow$ in both directions,
\begin{align}
\flow - \capacityflow &\le 0 \resultsin{\muupperflow} \Forall{\ell,t} 
\label[constraint]{eq:upper_flow_capacity_constraint} \\
- \flow - \capacityflow &\le 0 \resultsin{\mulowerflow} \Forall{\ell,t} 
\label[constraint]{eq:lower_flow_capacity_constraint}
\end{align}
which yield the shadow prices $\muupperflow$ and $\mulowerflow$. Again, we use the result of \cite{brown_decreasing_2020} (for details see \cref{sec:zero_profit_flow}) which derives that over the whole time span, the investment in line $\ell$ is payed back by the shadow prices times the flow 
\begin{align}
\capexflow_\ell = \capitalpriceflow \capacityflow = \sum_{t} \left( \muupperflow - \mulowerflow \right)  \flow 
\label{eq:no_profit_capex_flow}
\end{align}
Again, we follow the scheme in \cref{eq:general_scheme} which  defines the CAPEX allocation as 
\begin{align}
    \allocatecapexflow &=  
   \left( \muupperflow - \mulowerflow\right) \, \allocateflow
   \label{eq:allocate_capexFlow_detailed}
\end{align}
The quantity 
\begin{align}
 \allocateflow =  \pdv{\flow}{\demand}\, \demand
\end{align}
can be interpreted as the flow that the demand at $\left(n,t\right)$ causes on line $\ell$.
% According to \cref{eq:allocate_state_constraint} it has to fulfill 
% \begin{align}
%  \flow = \sum_{n} \allocateflow
% \label{eq:allocate_flow_constraint}
% \end{align}
The  shadowprices $\muupperflow$ and $\mulowerflow$ again can be seen as a measure for necessity of transmission investments at $\ell$ at time $t$. Hence, the definition of $\allocatecapexflow$ states that consumers, which retrieve power flowing on congested lines, yielding a bound \cref{eq:upper_flow_capacity_constraint} or \eqref{eq:lower_flow_capacity_constraint}, pay compensations for the resulting investments at $\ell$. Again the sum of all CAPEX payments to line $\ell$ equals the total CAPEX spent. This is seen when summing \cref{eq:allocate_capexFlow_detailed} over all buses and time steps and using \cref{eq:allocate_state_constraint,eq:no_profit_capex_flow}

@@

Given a fix price for emissions in CO$_2$ equivalents $\emissionprice$, the cost term for emission adds up to 
\begin{align}
 \emissioncost = \emissionprice \, \sum_s  \emission \, \generation
\end{align}
where $\emission$ denotes the emission factor in tonne-CO$_2$ per \megawatthour for generator $s$.
We use the same set of equation as given through \cref{eq:general_scheme,eq:opex_allocation} but update the cost factor to $\costfactor \Rightarrow \emissionprice \, \emission$. For consumers at bus $n$ at time $t$ this leads to the emission cost of 
\begin{align}
 \allocateemissioncost = \emissionprice \, \emission \, \allocategeneration
\end{align}
associated with generator $s$.

@@

The presented cost allocation suits for any type of topology and network setup. But so far, the question of how $\allocatestate$ for generators $s$, lines $\ell$ and storages $r$ are defined was left open. We recap that all rely on derivatives of $\generation$, $\flow$ and $\storagedispatch$  with respect to the nodal demand $\demand$ (see \cref{eq:allocate_state_constraint}). \\

Let $\nodalgeneration[m]$ denote the nodal power generation which combines the power production of all producing assets, in this case generators $S$ and storages $R$, at node $n$ and time $t$. It is given by 
\begin{align}
    \nodalgeneration[m] = \sum_{i \in \{S, R\}} \incidenceasset[m] \, \state
\end{align}
with $\incidenceasset[m]$ being 1 if asset $i$ is attached to bus $m$ and zero otherwise. Further, let $\allocatepeer$ collect the power produced by assets at node $m$ and consumed at $n$, given by 
\begin{align}
    \allocatepeer = \sum_{i \in \{S,R\}} \incidenceasset[m] \, \allocatestate  
    % \nodalgeneration[m] = \sum_s \incidencegenerator[m] \, \generation + \sum_r \incidencestorage[m] \storagedispatch \Forall{n}
\end{align}
Now, let $\ptdf$ denote Power Transfer Distribution Factors (PTDF) giving the changes in the flow on line $\ell$ for one unit (typically one MW) of net power production at bus $n$. The linear power flow equation can be written as 
\begin{align}
 \flow  = \sum_m \ptdf[m] \left( \nodalgeneration[m] - \demand[m] \right)  
\end{align}
Note that for transport models or mixed AC-DC networks, $\ptdf$ can be artificially calculated using the formulation presented in \cite{hofmann_flow_2020-1}.
Taking the derivative with respect to the demand, 
\begin{align}
 \allocateflow = \pdv{\flow}{\demand} \demand = \sum_m \ptdf[m] \left( \allocatepeer  - \delta_{n,m} \demand \right) 
 \label{eq:allocate_peer_to_allocate_flow},
\end{align}
shows that $\allocateflow$ is fully determined through the peer-to-peer allocation $\allocatepeer$. In other words, we only need to know how much power produced at node $m$ is consumed at node $n$ in order to derive the allocated flow $\allocateflow$. Further we can breakdown $\allocatepeer$ to $\allocategeneration$ for generators and $\allocatestoragedispatch$ for storages proportionally to their contribution to the nodal generation $\nodalgeneration[m]$. Unfortunately, the solution for $\allocatepeer$ is non-unique and requires further assumptions. Established flow allocation schemes approach this problem from different directions. Principally two options exist \textit{what} is allocated 
% 
\begin{enumerate}
\item gross power injections \label{gross}
\item net power injections \label{net}
\end{enumerate}
% 
Further it is important \textit{what assumptions} define the allocation, \ie what method is used to define the pairs of sources and sinks. The three suitable approaches we present here are
% 
\begin{enumerate}[label=\alph*., ref=\alph*]
\item Equivalent Bilateral Exchanges (EBE) \cite{galiana_transmission_2003} which assumes
that every producer supplies every consumer proportional to its share in the total consumption. \label{ebe} 
\item Average Participation (AP) \cite{bialek_tracing_1996,achayuthakan_electricity_2010-1} which traces the flow from producer to consumer following the law of proportional sharing. \label{ap}
\item Flow Based Market Coupling (FBMC) which uses zonal PTDF for allocating power within predefined regions. The interregional exchange is only allocating net power decifit or excess of the regions. \label{fbmc}
\end{enumerate}
% 
We show the mathematical formulation for all combinations \ref{ebe}\ref{gross} - \ref{fbmc}\ref{net} in \cref{sec:gross_ebe,sec:net_ebe,sec:gross_ap,sec:net_ap}.
% 
Principally, type \ref{net} leads to less P2P trades then type \ref{gross} as power from a bus $m$ with $\nodalgeneration[m] \le \nodaldemand[m]$ is not assigned to other buses, only to $m$. 
Further, as literature has often pointed out, the EBE principal \ref{ebe} does not suit for large networks where remote buses would interconnect in the same way as buses in close vicinity \cite{gil_multiarea_2005}. The AP based type \ref{ap} tackles this problem by restricting P2P trades to those which are traceable when applying the proportional sharing principal. Therefore $\allocategeneration$ denotes that part of power produced by bus $m$ which, when only following in the direction of $\flow$, ends up at bus $n$. Type \ref{fbmc} further allows to control the regions or market zones which are netted out in a first step. If in a region $R$ the generation undercuts the demand, $\sum_{n \in R} \nodalgeneration \le \sum_{n \in R} \demand$, none of the inner-regional generation is assigned to other regions.

@@

In the following, we illustrate the cost allocation and explain its functionality by taking the example of a two bus system with one optimized time step, shown in \cref{fig:example_network}. \\
% 
% 
\begin{figure*}[t]
\centering
\includegraphics[width=\linewidth]{example_network.png}
\caption{Illustrative example of a 2-bus network with one optimized time step. Fixed prices and constraining values are given in the left box for each bus and the transmission line. Optimized values are given in the right boxes. Bus~1 has a cheaper operational price $o$, capital prices are the same for both. As both generator capacities are constraint to 100~MW, the optimization also deploys the generator at bus~2. The resulting electricity prices $\lambda$ are then a composition of all prices for operation and capital investments. \vspace{-10pt}}
\label{fig:example_network}
\end{figure*}
% 
% 
The two buses are connected via one transmission line, each has one generator. Whereas the generator at bus~1 has an operational price of 50 \euro/\megawatthour, the generator at bus~2 has a higher operational price of 200~\euro/\megawatthour. For both the CAPEX rate is set to 500~\euro/MW and the maxmimal capacity is limited to $\capacitygenerationupper$~=~100~MW. The transmission line has a CAPEX rate of 100~\euro/MW and no upper capacity limit. With a demand of 60~MW at bus~1 and 90~MW at bus~2, the optimization expands the cheaper generator at bus~1 to its full limit of 100~MW. The 40~MW excess power not consumed at bus~1, flow to bus~2 where the generator is built with only 50~MW.  
% 
\begin{figure}[h!]
    \begin{subfigure}[c]{\linewidth}
    \includegraphics[width=\linewidth]{example_allocation_bus1.png}
    \vspace{-40pt}
    \subcaption{}
    \label{fig:example_allocation_bus1}
    \end{subfigure}
    \begin{subfigure}[c]{\linewidth}
    \includegraphics[width=\linewidth]{example_allocation_bus2.png}
    \vspace{-40pt}
    \subcaption{}
    \label{fig:example_allocation_bus2}
    \end{subfigure}
    \caption{Power allocations for bus~1 (a) and bus~2 (b) of the example network in \cref{fig:example_network}. Bus~1 retrieves 40~MW from itself and 20~MW from bus~2. The latter in turn retrieves 60~MW from bus~1 and self supplies 30~MW.
    The sum of both net flows equals the resulting flow of $f_1=40$~MW. \vspace{-20pt}}
    \label{fig:example_allocation}
\end{figure}
% 
% 
% 
\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{example_payoff.png}
\caption{Full P2P cost allocation for the example setup shown in \cref{fig:example_network}. The payments are derived on the basis of \cref{eq:allocate_opexGeneration_detailed,eq:allocate_capexGeneration_detailed,eq:allocate_capexFlow_detailed}. Consumers at bus $n$ have to pay each generator proportional to their consumption. As we only consider one time step the proportionality applies for OPEX $\allocateopex$ and CAPEX $\mathcal{C}_n$. As bus~1 induces a relieving flow an line~1 and therefore ``prevents`` further tranmission expansion, it is rewarded proportional to the relief.}
\label{fig:example_payoff}
\end{figure}
% 
% 
\subsubsection*{Allocating Gross Power using EBE (\ref{gross}\ref{ebe}) }

\Cref{fig:example_allocation} shows the allocated transactions on basis of equvalently traded gross power (type \ref{gross}\ref{ebe}) for both buses 1~\&~2 separately. The resulting P2P payments are given in \cref{fig:example_payoff}.
The upper graph \cref{fig:example_allocation_bus1} shows that $A_{1,1}=40$~MW at bus~1 are self-sustained. Consumers at bus~1 consequently pay 2k~\euro~OPEX and 22k~\euro~CAPEX to the generator at bus~1. The remaining 20~MW come from bus~2 and induce a subflow on line 1 of $A_{\ell = 1, 1} = -20$. As this flow is in contrary direction to the total flow, it is relieving the transmission system. This translates to a congestion reward for consumers at bus~1 of $c_{\ell=1} A_{\ell = 1, 1}$~=~2k~\euro\, which is exactly the cost that had to be spent on the transmission system if bus~1 didn't induce a relieving flow, see again \cref{fig:example_payoff}.

@@

\begin{align}
    \capitalpricegeneration \capacitygeneration - \remainingcost_s = \sum_{n,t} \allocatecapexgeneration \Forall{s \in S}
\end{align}

Doing likewise with a subset $L$ of transmission lines,
\begin{align}
\capacityflow \ge \capacityflowLower \resultsin{\mulowerflownom} \hpad \forall{\ell \in \textit{L}} \\
\capacityflow \le \capacityflowUpper \resultsin{\muupperflownom} \hpad \forall{\ell \in \textit{L}}
\end{align}
results in a shifted zero-profit rule for transmission line in the form 
\begin{align}
    \capitalpriceflow \, \capacityflow - \remainingcost_\ell = \sum_{n,t}\allocatecapexflow \Forall{\ell \in L}
\end{align}
where $\remainingcost_\ell$ is given by 
\begin{align}
    \remainingcost_\ell = \left(\mulowerflownom - \muupperflownom \right) \, \capacityflow
    \Forall{\ell \in L}
\end{align}
